Method

### 2025-11-26 GroupPolarizationStatmod/paper/method.qmd
updated method, started Discussion standalone

All bins have a width of 1 in latent opinion space, except for the two most
extreme bins at opposite ends of the sentiment spectrum---this works because bin
values are always separated by 1. All continuous latent opinion values
$\omega < b_1 + \frac{1}{2} = \theta_1$ get mapped to the ordinal measurement
value $b_1$---so, $b_1$'s lower threshold is $\theta_0 = -\infty$. Similarly,
all opinions $\omega > b_B - 0.5 = \theta_B$ are reported as the maximum bin
value, $b_B$, with an effective upper bound of $\theta_{B+1} = +\infty$. We can
define all thresholds compactly as [@Liddell2018] $$
\theta_j = \begin{cases}
  -\infty         & \text{ if } j=0 \\
  \infty          & \text{ if } j=B \\
  b_j + \frac{1}{2} & \text{ otherwise.}
\end{cases}
$$

@@

When we plug in the bin values and copmuted $f_\mathrm{pre}$ values into the
simulated average pre-discussion opinion definition (@eq-sim-ave-observed), $$
\bar{o}_\mathrm{pre} = -3 \cdot 0.38~~+-2\cdot 0.09~~+\ldots+3\cdot0.19 = -0.6
$$ $$
\bar{o}_\mathrm{post} = -3 \cdot 0.25~~+-2\cdot 0.19~~+\ldots+3\cdot0.03 = -1.04
$$ So, we find that for the `Moscovici1969_Americans`, our null-polarization
model appears polarized, with the exact same average pre- and post-discussion
opinions reported by Moscovici and Zavalloni (1969), and the same magnitude of
radicalization (CHECK THE PRESENTATION OF ALL THREE ABOVE) $$
g(\nu, e) = \bar{o}_\mathrm{post} - \bar{o}_\mathrm{pre} = -1.04 - (-0.6) = - 0.4.
$$

@@

To evaluate an experimental design's ability to properly identify null effects,
we need to simulate participant responses assuming the mere agreement null model
that invalidated the replication, if one was found. We simulate data from a
single group polarization experiment by sampling participants' ordinal opinions
at pre-/post-discussion from the pre- and post-discussion opinion distributions
specified by the null model that invalidated the replication in that experiment,
$\nu_e$. If there are $N_e$ participants in experiment $e$ then our simulated
data sets contained $2N_e$ rows. We recorded $N_e$ in the web app as metadata
along with data about average opinions and opinion shifts for each experimental
condition across the corpus.

It can be helpful to

@@

In the final analysis step we calculate what significance threshold, $d^*$,
would limit the false positive rate to 0.05 or less.

There are many possible null models we could choose.

what significance threshold would have limited inverting this procedure to find
what significance value $d^*$ is necessary to achieve a low family-wise error
rate $\alpha_e = 0.05$ for experimental condition, $e$.

To do this we first calculate $\alpha_e(d^*)$ for a sufficiently wide range of
$d^*$. We then find the lowest $d^*$ for which $\alpha_e(d^*) \leq 0.05$.

>>>>

All bins have a width of 1 in latent opinion space, except for the two most
extreme bins at opposite ends of the sentiment spectrum---this works because bin
values are always separated by 1. All continuous latent opinion values
$\omega < b_1 + \frac{1}{2} = \theta_1$ get mapped to the ordinal measurement
value $b_1$---so, $b_1$'s lower threshold is $\theta_0 = -\infty$. Similarly,
all opinions $\omega > b_B - 0.5 = \theta_B$ are reported as the maximum bin
value, $b_B$, with an effective upper bound of $\theta_{B+1} = +\infty$. We can
define all thresholds compactly as [@Liddell2018] $$
\theta_j = \begin{cases}
  -\infty         & \text{ if } j=0 \\
  \infty          & \text{ if } j=B \\
  b_j + \frac{1}{2} & \text{ otherwise.}
\end{cases}
$$ {#eq-bin-thresholds}

@@

When we plug in the bin values and copmuted $f_\mathrm{pre}$ values into the
simulated average pre-discussion opinion definition (@eq-sim-ave-observed), we
get $$
\bar{o}_\mathrm{pre} = -3 \cdot 0.38~~+-2\cdot 0.09~~+\ldots+3\cdot0.19 = -0.6
$$ and $$
\bar{o}_\mathrm{post} = -3 \cdot 0.25~~+-2\cdot 0.19~~+\ldots+3\cdot0.03 = -1.04.
$$ So, we find that for the `Moscovici1969_Americans`, our null-polarization
model appears polarized, with the exact same average pre- and post-discussion
opinions reported by Moscovici and Zavalloni (1969), and the same magnitude of
radicalization: $$
g(\nu, e) = \bar{o}_\mathrm{post} - \bar{o}_\mathrm{pre} = -1.04 - (-0.6) = - 0.4.
$$

@@

To evaluate an experimental design's ability to properly identify null effects,
we need to simulate participant responses assuming the mere agreement null model
that invalidated the replication ($\nu_e$). We use the stochastic model for this
simulation stage, as it correctly models the noise inherent in individual
participant measurement and response. The ordinal, measured opinions are
generated by executing the following steps for $N_e$ simulated participants
across $N_T$ independent trials:

1.  **Latent Opinion Sampling:** For each participant $i$ and time
    $t \in \{\text{pre, post}\}$, we draw a continuous latent opinion
    $\omega_{i,t}$ from the corresponding normal distribution defined by the
    null model $\nu_e$: $$\omega_{i,t} \sim \mathcal{N}(\mu, \sigma_t)$$ where
    $\mu$ is the constant latent mean and $\sigma_t$ is the time-specific
    standard deviation derived from the mere agreement null model $\nu_e$.
2.  **Measurement Simulation (Ordinal Binning):** Each continuous latent opinion
    $\omega_{i,t}$ is mapped onto the experiment's ordinal scale (e.g., $-3$ to
    $+3$). This mapping is determined by the fixed measurement thresholds,
    $\theta_j$, established in the *Expectation Model* subsection
    (@eq-bin-thresholds). This process simulates the clipping artifact that
    generates spurious polarization in the observable data.

This results in a simulated dataset of $2N_e$ rows (two time steps per
participant) for a single trial, consisting of ordinal opinion values that are
then passed to the statistical inference pipeline.

@@

In the final stage of our analysis, we pivot from evaluating the outcome of
published designs to *specifying the statistical requirements for future
designs*. The primary goal of this specification is to invert the standard
question of statistical rigor: instead of asking if an effect is 'significant'
given a fixed effect-size threshold (e.g., $d^{*} = 0.5$), we ask what
significance threshold is necessary to limit the false positive rate to a
desired maximum, such as $5\%$ (or $\alpha = 0.05$).

We leverage the evaluation framework developed in Part 2, based on the mere
agreement null model $\nu_e$ identified for each experimental condition $e$. Our
objective is to calculate the function $\alpha_e(d^*)$ across a sufficient range
of $d^*$ values. This allows us to determine the lowest $d^*$ that satisfies the
design requirement $\alpha_e(d^*) \leq 0.05$.

The function $\alpha_e(d^*)$ is the simulated false positive rate as a function
of the significance threshold, $d^*$, defined as: $$
\alpha_e(d^*) = \frac{1}{N_T} \sum_i^{N_T} \Theta(d_{ei}; d^*),
$$ where $N_T$ is the number of null trials, $d_{ei}$ is the calculated effect
size for trial $i$, and $\Theta$ is the indicator function that flags trials
where the effect size $d_{ei}$ exceeds the significance threshold $d^*$. In the
interest of giving the most generous best-case scenario possible, we only count
positive effect sizes because only these indicate increased extremism over time.
Negative effect sizes indicate group opinions became more moderate—such
observations do not support the group polarization hypothesis that opinions
become more extreme following discussion.

Because our system is automated, and the code publicly available, it can be
adapted and used to design future group polarization experiments with
empirically-grounded statistical rigor. Experimenters can specify a desired
maximum false positive rate, such as $\alpha = 0.05$, and our system will
calculate the minimum effect size threshold $d^*$ that must be exceeded to
achieve that level of rigor under the mere agreement null model identified for
their experimental design. This prescriptive step transforms the simulation from
a post-hoc critique into a forward-looking design tool, giving experimenters an
empirically-derived minimum effect size they must exceed to maintain statistical
confidence in their results.

# Analysis Outline and Summary
